Off the beaten path: \\ Gaia reveals GD-1 stars outside of the main stream

### 2018-05-01 paper/gd1.tex
ab trims

\begin{abstract}\noindent % trust me
Tidally-disrupted globular clusters are transformed into thin, dynamically-cold
streams of stars that are extremely valuable tracers of the large- and
small-scale distribution of mass in the Galaxy.
Most stellar streams discovered around the Milky Way reside in the Galactic halo
and are therefore primarily sensitive to the dark matter distribution.
% The longest and most prominent thin stream in the Milky Way --- the GD-1 stream --- has been used
% to constrain the shape of the inner halo, and is a promising place to search for
% signatures of dark matter subhalo interactions.
Using the exquisite data from the \gaia\ second data release combined with
\pans\ photometry,  we present a sample of highly-probable members of the
longest cold stream, GD-1, that (1) extend the apparent length of the stream by
$20^\circ$, (2) reveal a plausible location for the progenitor, and (3) indicate
the existence of stream members off of the main stream track.
% selected to have
% small parallaxes and retrograde proper motions using data from the \gaia\
% mission, and \pans\ photometry consistent with an old and metal-poor population
% at a distance of $\sim10~\kpc$.
% This selection extends the stream by $20^\circ$ in apparent length, reveals a
% plausible location of the progenitor, and shows significant density variations
% along the stream in unprecedented detail.
This new view of the stream reveals significant density variations
along the stream in unprecedented detail.
% In addition to confirming several prominent gaps detected in photometric maps of
% GD-1, for the first time we find evidence of stream members off of the main
% track.
These features could be an indication of complex internal kinematics within the
progenitor unlike any of the known globular clusters, or a result of a stream
interaction with a massive perturber.
% Detailed modeling of the positional and velocity offsets
% or an indication of complex internal kinematics within the
% stream progenitor.
% Velocity offset between these features and the stream is an order of magnitude larger than the variance in velocity fields of known globular cluster, thus favoring the encounter origin.
% If this is confirmed, off-stream features will enable a robust estimate of the perturber's mass.
% If caused by a subhalo interaction, we estimate that the detected underdensity
% and the associated spur could have been induced by a perturbation of a
% $\sim10^7\,\rm M_\odot$(?) object $\sim 0.5\,$Gyr ago.
\end{abstract}

@@

Stellar streams are formed during the tidal disruption of stellar systems by the
gravitational field of their host galaxy.
The phase-space density and mean track of stars in streams therefore encode
information about the underlying distribution and shape of mass on galactic
scales (e.g., \citealt{Johnston:1999, Bonaca:2018}).
% Most stellar streams are found in the halos of galaxies and are therefore
% important tracers of the large-scale distribution of dark matter in galactic
% halos.
% Long, thin stellar streams are also excellent laboratories for studying
% small-scale structures in the mass distribution:
% thin streams can remain coherent for tens of orbital periods, depending in
% detail on the symmetries of the underlying galactic mass distribution (e.g.,
% \citealt{Erkal:2016a}).
Dynamically-cold stellar streams formed from disrupted globular clusters can
also retain imprints on their phase-space density from encounters with a massive
perturber for long times \citep[e.g.,][]{Yoon:2011}.
% Encounters between massive perturbers and stream stars imprint variations in the
% phase-space density that can persist and remain visible for close to their total
% survival time (e.g., \citealt{Yoon:2011}).
In this sense, streams represent one of the most promising directions for
testing the existence of small-scale dark matter sub-halos, predicted by
standard $\Lambda$ Cold Dark Matter ($\Lambda$CDM) theory
(\citealt{Erkal:2015, Sanders:2016, Bovy:2017}).

Over 30 candidate stellar streams have been discovered in the Milky Way \citep[see][for a recent review]{Grillmair:2016, Newberg:2016}.
% thanks to large-area, multi-band photometric surveys (especially
% the Sloan Digital Sky Survey, SDSS; \citealt{York:2000}).
% The Milky Way streams have a wide range of properties such as length, velocity
% dispersion, and density (e.g., \citealt{Odenkirchen:2001, Grillmair:2006,
% Grillmair:2006b, Belokurov:2006, Belokurov:2007, Bonaca:2012, Shipp:2018}; see
% also the recent review \citealt{Grillmair:2016, Newberg:2016}).
% , and the majority
% have no obvious, bound progenitor systems;
% of the thin, likely globular-cluster-origin streams, only two have clear
% progenitors (NGC 5466 and Palomar 5).
% Most streams were first detected as a linear overdensity in stellar counts filtered to preferentially select faint blue stars, most likely members of a former Galactic satellite.
The GD-1 stream is the most prominent among the thin streams, discovered as an
overdensity of faint blue stars in the SDSS photometry \citep{Grillmair:2006}.
GD-1 spans $\sim 60^\circ$ on the sky, which, at its distance ($\sim
7$--$10~\textrm{kpc}$) translates to a physical length of $\sim 15~\kpc$.
No remnant progenitor for the stream has been found, but its width ($\sigma
\approx 12'$), metallicity ($\feh \approx -1.4$), and estimated stellar mass ($M
\approx 2 \times 10^4~\msun$) imply that GD-1 is a disrupted globular cluster
\citep{Koposov:2010}.

@@

Its implied old dynamical age, combined with its orbital properties, makes GD-1
a prime stream to also study the small scale structure of dark matter.
While density variations in streams can be induced from interactions with giant
molecular clouds \citep{Amorisco:2016} or resonant encounters with the bar
\citep{Pearson:2017}, these baryonic effects are unlikely to affect GD-1 because
of its large pericentric distance and retrograde orbit with respect to the
Galactic bar.
Therefore, observed density variations in the GD-1 stream could instead be an
indication of past interactions with dark matter subhalos
\citep[e.g.,][]{Ngan:2014}.
By estimating a stream age of $\approx 4~\textrm{Gyr}$, assuming a progenitor
mass of $10^5~\msun$, and by assuming properties of the subhalo population from
% dark matter-only simulations \citep[scaled to the Milky Way halo mass and
% reduced in number by a factor of 3 to account for disruption by the Galactic
% disk]{Springel:2008, Diemand:2008}, \citet{Erkal:2016} predict that GD-1 could
dark matter-only simulations \citep{Springel:2008, Diemand:2008},
\citet{Erkal:2016} predict that GD-1 could have up to one significant, wide
($\approx 5$--$7^\circ$) gap caused by an interaction with a
$10^6$--$10^7~\msun$ subhalo.
A definitive ruling on the presence of dark matter subhalos in this mass regime
would directly inform the nature of dark matter \citep[e.g.][]{Bullock:2017}.

@@

We retrieve \gaia\ data along the previously-identified track of the GD-1 stream
from the \gaia\ science archive\footnote{\url{https://gea.esac.esa.int/}} by
selecting sources with small parallax, $\varpi < 1~\textrm{mas}$, and select
batches of stars that have small latitude in the heliocentric GD-1 coordinate
system, $(\phi_1, \phi_2)$, defined in \cite{Koposov:2010}.
We convert the sky coordinates and proper motions of the parallax-selected stars
to the GD-1 stream system, and correct the proper motions for solar reflex
motion by assuming that stars at a given stream longitude, $\phi_1$, have a
distance given by $d(\phi_1) = (0.05 \, \phi_1 + 10)~\textrm{kpc}$;
we assume a solar velocity $\bs{v}_\odot = (11.1, 232.24, 7.25)~\kms$
(\citealt{Schonrich:2010, Bovy:2015}).
\figurename~\ref{fig:selection} (top right) shows stars with $|\phi_2| <
1^\circ$ in solar-reflex-corrected proper motion components in the GD-1 system,
$\mu_{\phi_1, \odot}$ and $\mu_{\phi_2, \odot}$.
GD-1 members are visible as the overdensity around $(\mu_{\phi_1, \odot},
\mu_{\phi_2, \odot}) \approx (-8, 0)~\masyr$.

@@

To improve the contrast of the stream over the background, we cross-match the
sample to the \pans\ photometric catalog.% and use the photometry to clean the sample.
\figurename~\ref{fig:selection} (bottom right) shows the proper-motion-selected
candidate stream stars within $\left|\phi_2\right| < 1^\circ$ in a \pans\
color-magnitude diagram (CMD) de-reddened following \citet{Schlafly:2011}.
The over-plotted isochrone (red line) is from the MESA Isochrones \& Stellar
Tracks (MIST; \citealt{Dotter:2016, Choi:2016, Paxton:2011}) and represents a
$12~\textrm{Gyr}$ old population with $\feh = -1.35$ at a distance of
$7.8~\kpc$.
We use this isochrone to generate a polygonal selection in de-reddened $g-i$
color and apparent $g$-band magnitude, as visualized by the shaded region in the
bottom right panel of \figurename~\ref{fig:selection}.

\figurename~\ref{fig:selection} (bottom left) shows the final sample of
candidate stream members after selecting on both proper motion and photometry.
The GD-1 stream is clearly visible in the point density of \emph{individual}
stars (the positions are not binned); this is the most pure view of the stream
to date.
This increased contrast shows that the stream extends at least another
$20^\circ$ to negative longitudes.
Furthermore, the stream reaches its highest surface density where it is the
narrowest ($\phi_1\sim-13^\circ$), which may be the location of its elusive and
fully-disrupted progenitor.
Several of the under-densities and gaps hinted at from photometric selection
(\citealt{Carlberg:2013, DeBoer:2018}) appear as striking features in this
significantly cleaner map of the GD-1 stream.
% Finally, at least two new features are visible near the stream: (1) ``the
% spur,'' stars above the main stream track (in $\phi_2$) between $-40^\circ
% \lesssim \phi_1 \lesssim -30^\circ$, and (2) ``the blob,'' stars below the main
% stream track between $-20^\circ \lesssim \phi_1 \lesssim -10^\circ$.
% We discuss each of these in more detail in \sectionname
% s~\ref{sec:results}--\ref{sec:discussion} below.
Finally, at least two new features are visible near the stream: (1) ``the
spur,'' stars above the main stream track (in $\phi_2$), and (2) ``the blob,''
stars below the main stream track, both highlighted in
\figurename~\ref{fig:selection}.
We discuss each of these in more detail in \S\,
\ref{sec:results}--\ref{sec:discussion} below.

@@

We extract high-confidence stream stars from the proper motion and CMD selected
sample by further selecting stars close to the main stream track.
We use these main stream track stars to study the global properties of the
stream (\S\,\ref{sec:res_global}), and find interesting substructures off of the main stream track (\S\,\ref{sec:res_gap}).

\subsection{Global properties}
\label{sec:res_global}

@@

We extract properties of the stream from the region defined above as a function
of $\phi_1$ by computing median stream properties in overlapping windows:
we set the window size to $3^\circ$ in $\phi_1$, and shift the window center by
$1^\circ$ in longitude for each subsequent measurement.
Panels (c)--(e) in \figurename~\ref{fig:track-and-model} show the
background-subtracted stream surface density and median proper motion components
extracted in this way (dark lines), with the associated empirical scatters
(shaded regions).
We estimate the local background surface density as an average of areas north
and south of the bounded stream region.
The median proper motion profiles along the stream (panels (d)--(e)) are
consistent with previous mean proper motion measurements
(\citealt{Koposov:2010}), but here the proper motion gradient is resolved for
individual stars (see, e.g., elongated overdensity in upper right panel of
\figurename~\ref{fig:selection}).

@@

Clear density variations, also apparent in the 2D positions of the sources,
manifest as sharp features in the surface density estimates along the stream.
These are not a result of variable dust extinction or completeness variations
from the \gaia\ scanning pattern.
First, the visible portions of the GD-1 stream are located at high Galactic
latitudes ($b > 20^\circ$).
%, and we therefore do not expect significant dust
% extinction or variations in extinction within the stream region.
% The median $V$-band extinction over the stream track is
% $\textrm{med}\left(A_V\right) = 0.04~\textrm{mag}$ with a dispersion
% $\sigma_{A_V} \approx 0.03~\textrm{mag}$, as computed from the SFD extinction
% map (\citealt{Schlegel:1998}).
% As the stream approaches the Galactic disk ($\phi_1 < -60^\circ$),
% dust extinction becomes appreciable, and the stream stars are harder to
% select apart from the background.
Second, 99\% of the stars in our sample have at least 8 \gaia\ visibility
periods (median and root-variance of 14 and 3, respectively), promising a robust
estimate of the proper motions.
% The variance in the number density of faint stars originating from the \gaia\
% scanning pattern is only $\approx 1~\textrm{deg}^{-2}$, as estimated using stars
% with proper motions $-9 < \mu_{\phi_1, \odot} < -4.5~\masyr$ and $1 <
% \mu_{\phi_2, \odot} < 4~\masyr$, far smaller than the density variations in the
% stream.
% The relative ratios between different over- and under-densities might be affected by the \gaia\ completeness at the faint end, but they

The observed density variations are therefore likely real morphological changes
along the GD-1 stream, as have been suggested before photometric selection alone
(\citealt{Carlberg:2013, DeBoer:2018}).
The deep under-densities we see from the kinematically-cleaned sample correspond
to previously reported gaps in the stream.
There is also a well-defined, sharp overdensity in the surface density profile
close to $\phi_1 \approx -13^\circ$ with roughly symmetric under-densities on
either side:
this is suggestive of a progenitor system in the final stages of dissolution
(\citealt{Balbinot:2018}).
This region is also the thinnest part of the stream, as would be expected for
recently stripped material.
% that has not yet phase-mixed in multiple dimensions.

@@

To compute initial conditions for the stream model, we fit an orbit to the
observed properties of the stream in a fixed Milky Way model consisting of a
disk (\citealt{Miyamoto:1975}), a bulge (\citealt{Hernquist:1990}), and a
spherical dark matter halo (\citealt{Navarro:1996}).
We set the disk scale-length parameters to match the disk model found in
\citet{Bovy:2015}; other parameters are set to produce an approximately flat
disk rotation curve with a circular velocity at the solar circle,
$v_{\textrm{circ}, \odot} \approx 230~\kms$.
%use L-BFGS-B optimization as implemented in the \package{scipy}
% \texttt{Python} package to
We maximize the likelihood of the orbit given sky track and proper motions from
this work, with mean distance and giant star radial velocities from
\cite{Koposov:2010}.
We do not vary the parameters of the Milky Way model in the fit because we are
only interested in producing a comparison stream model. % of the stream properties.

\figurename~\ref{fig:orbitfit} shows the best-fitting orbit from the procedure
outlined above.
The GD-1 stream crosses the midplane at large radius, far from the bulk of the
stellar density:
the grey line in the left panel shows the radius that contains $\approx 90\%$ of
the mass, assuming an exponential radial density profile with a scale radius
$\rm R_d = 3~\kpc$ (an intermediate value of recent measurements, e.g.,
\citealt{Bovy:2012}).

@@

To initialize the model stream generation, we assume that the present day
position of the progenitor system is at $\phi_1 = -13.5^\circ$, the peak of the
observed surface density profile.
We numerically integrate the backwards from this location for $4~\textrm{Gyr}$,
then forward-generate the model stream, and release stream particles at each
timestep.
We assume an initial progenitor mass of $M=10^5~\msun$, and linearly decrease
the mass of the progenitor until present day, where we assume $M = 0~\msun$
(i.e. fully disrupted).
Stream properties computed from this resulting model stream are plotted in
\figurename~\ref{fig:track-and-model} as dashed (orange) lines.
The model stream qualitatively matches the stream track in all phase-space
dimensions, but has a much smoother surface density profile along the extent of
the stream.
We therefore don't expect that the observed surface density variations could be
caused by, e.g., epicyclic overdensities (\citealt{Kupper:2012}).

>>>>

\begin{abstract}\noindent % trust me
Using the exquisite data from the \gaia\ second data release combined with
\pans\ photometry,  we present a sample of highly-probable members of the
longest cold stream in the Milky Way -- GD-1.
The resulting map of GD-1: (1) extends the apparent length of the stream by
$20^\circ$, (2) reveals a plausible location for the progenitor, (3) detects 
significant density variations along the stream, and (4) indicates
the existence of stream members off the main stream track.
% Tidally-disrupted globular clusters are transformed into thin, dynamically-cold
% streams of stars that are extremely valuable tracers of the large- and
% small-scale distribution of mass in the Galaxy.
% Most stellar streams discovered around the Milky Way reside in the Galactic halo
% and are therefore primarily sensitive to the dark matter distribution.
% The longest and most prominent thin stream in the Milky Way --- the GD-1 stream --- has been used
% to constrain the shape of the inner halo, and is a promising place to search for
% signatures of dark matter subhalo interactions.
% selected to have
% small parallaxes and retrograde proper motions using data from the \gaia\
% mission, and \pans\ photometry consistent with an old and metal-poor population
% at a distance of $\sim10~\kpc$.
% This selection extends the stream by $20^\circ$ in apparent length, reveals a
% plausible location of the progenitor, and shows significant density variations
% along the stream in unprecedented detail.
% This new view of the stream reveals significant density variations
% along the stream in unprecedented detail.
% In addition to confirming several prominent gaps detected in photometric maps of
% GD-1, for the first time we find evidence of stream members off of the main
% track.
These features could be an indication of complex internal kinematics within the
progenitor unlike any of the known globular clusters, or a result of a stream
interaction with a massive perturber.
% Detailed modeling of the positional and velocity offsets
% or an indication of complex internal kinematics within the
% stream progenitor.
% Velocity offset between these features and the stream is an order of magnitude larger than the variance in velocity fields of known globular cluster, thus favoring the encounter origin.
% If this is confirmed, off-stream features will enable a robust estimate of the perturber's mass.
% If caused by a subhalo interaction, we estimate that the detected underdensity
% and the associated spur could have been induced by a perturbation of a
% $\sim10^7\,\rm M_\odot$(?) object $\sim 0.5\,$Gyr ago.
\end{abstract}

@@

Stellar streams are formed during the tidal disruption of stellar systems by the
gravitational field of their host galaxy.
The phase-space density and mean track of stars in streams therefore encode
information about the underlying distribution mass on galactic
scales (e.g., \citealt{Johnston:1999, Bonaca:2018}).
% Most stellar streams are found in the halos of galaxies and are therefore
% important tracers of the large-scale distribution of dark matter in galactic
% halos.
% Long, thin stellar streams are also excellent laboratories for studying
% small-scale structures in the mass distribution:
% thin streams can remain coherent for tens of orbital periods, depending in
% detail on the symmetries of the underlying galactic mass distribution (e.g.,
% \citealt{Erkal:2016a}).
Dynamically-cold stellar streams formed from disrupted globular clusters can
also retain imprints on their phase-space density from encounters with a massive
perturber for long times \citep[e.g.,][]{Yoon:2011}.
% Encounters between massive perturbers and stream stars imprint variations in the
% phase-space density that can persist and remain visible for close to their total
% survival time (e.g., \citealt{Yoon:2011}).
In this sense, streams represent one of the most promising directions for
testing the existence of small-scale dark matter sub-halos, predicted by
standard $\Lambda$ Cold Dark Matter ($\Lambda$CDM) theory
(\citealt{Erkal:2015, Sanders:2016, Bovy:2017}).

Over 30 candidate stellar streams have been discovered in the Milky Way \citep[see][for a recent review]{Grillmair:2016, Newberg:2016}.
% thanks to large-area, multi-band photometric surveys (especially
% the Sloan Digital Sky Survey, SDSS; \citealt{York:2000}).
% The Milky Way streams have a wide range of properties such as length, velocity
% dispersion, and density (e.g., \citealt{Odenkirchen:2001, Grillmair:2006,
% Grillmair:2006b, Belokurov:2006, Belokurov:2007, Bonaca:2012, Shipp:2018}; see
% also the recent review \citealt{Grillmair:2016, Newberg:2016}).
% , and the majority
% have no obvious, bound progenitor systems;
% of the thin, likely globular-cluster-origin streams, only two have clear
% progenitors (NGC 5466 and Palomar 5).
% Most streams were first detected as a linear overdensity in stellar counts filtered to preferentially select faint blue stars, most likely members of a former Galactic satellite.
The GD-1 stream is the most prominent among the thin streams, discovered as an
overdensity of faint blue stars in the SDSS photometry \citep{Grillmair:2006}.
Initially detected to span $\sim 60^\circ$ on the sky, the physical length of GD-1 at a distance of $\sim
7$--$10~\textrm{kpc}$ is at least $\sim 15~\kpc$.
No remnant progenitor for the stream has been found, but its width ($\sigma
\approx 12'$), metallicity ($\feh \approx -1.4$), and estimated stellar mass ($M
\approx 2 \times 10^4~\msun$) imply that GD-1 is a disrupted globular cluster
\citep{Koposov:2010}.

@@

Its implied old dynamical age, combined with its orbital properties, makes GD-1
a prime stream to also study the small scale structure of dark matter.
While density variations in streams can be induced from interactions with giant
molecular clouds \citep{Amorisco:2016} or resonant encounters with the bar
\citep{Pearson:2017}, these baryonic effects are unlikely to affect GD-1 because
of its large pericentric distance and retrograde orbit with respect to the
Galactic bar.
Therefore, observed density variations in the GD-1 stream could instead be an
indication of past interactions with dark matter subhalos
\citep[e.g.,][]{Ngan:2014}.
By estimating a stream age of $\approx 4~\textrm{Gyr}$, assuming a progenitor
mass of $10^5~\msun$, and by assuming properties of the subhalo population from
% dark matter-only simulations \citep[scaled to the Milky Way halo mass and
% reduced in number by a factor of 3 to account for disruption by the Galactic
% disk]{Springel:2008, Diemand:2008}, \citet{Erkal:2016} predict that GD-1 could
dark matter-only simulations \citep{Springel:2008, Diemand:2008},
\citet{Erkal:2016} predict that GD-1 could have up to one significant, wide
($\approx 5$--$7^\circ$) gap caused by an interaction with a
$10^6$--$10^7~\msun$ subhalo.
A definitive ruling on the presence of dark matter subhalos in this mass regime
would directly inform the nature of dark matter \citep[e.g.,][]{Bullock:2017}.

@@

We retrieve \gaia\ data along the previously-identified track of the GD-1 stream
from the \gaia\ science archive\footnote{\url{https://gea.esac.esa.int/}} by
selecting sources with small parallax, $\varpi < 1~\textrm{mas}$, and select
batches of stars that have small latitude in the heliocentric GD-1 coordinate
system, $(\phi_1, \phi_2)$, defined in \cite{Koposov:2010}.
We convert the equatorial sky coordinates and proper motions %of the parallax-selected stars
to the GD-1 stream system, and correct the proper motions for solar reflex
motion by assuming that stars at a given stream longitude, $\phi_1$, have a
distance given by $d(\phi_1) = (0.05 \, \phi_1 + 10)~\textrm{kpc}$;
we assume a solar velocity $\bs{v}_\odot = (11.1, 232.24, 7.25)~\kms$
(\citealt{Schonrich:2010, Bovy:2015}).
\figurename~\ref{fig:selection} (top right) shows stars with $|\phi_2| <
1^\circ$ in solar-reflex-corrected proper motion components in the GD-1 system,
$\mu_{\phi_1, \odot}$ and $\mu_{\phi_2, \odot}$.
GD-1 members are visible as the overdensity around $(\mu_{\phi_1, \odot},
\mu_{\phi_2, \odot}) \approx (-8, 0)~\masyr$.

@@

To improve the contrast of the stream over the background, we cross-match the
sample to the \pans\ photometric catalog. % and use the photometry to clean the sample.
\figurename~\ref{fig:selection} (bottom right) shows the proper-motion-selected
candidate stream stars within $\left|\phi_2\right| < 1^\circ$ in a \pans\
color-magnitude diagram (CMD) de-reddened following \citet{Schlafly:2011}.
GD-1 stellar population stands out from the Milky Way foreground, and is well matched by a 12\,Gyr, $\feh = -1.35$ MIST isochrone at a distance of $7.8\,\kpc$ \citep[red line in Figure~\ref{fig:selection}][]{Dotter:2016, Choi:2016, Paxton:2011}. 
% The over-plotted isochrone (red line) is from the MESA Isochrones \& Stellar
% Tracks (MIST; \citealt{Dotter:2016, Choi:2016, Paxton:2011}) and represents a
% $12~\textrm{Gyr}$ old population with $\feh = -1.35$ at a distance of
% $7.8~\kpc$.
We use this isochrone to generate a polygonal selection in de-reddened $g-i$
color and apparent $g$-band magnitude (shaded region in the
bottom right panel of \figurename~\ref{fig:selection}).

\figurename~\ref{fig:selection} (bottom left) shows the final sample of
candidate stream members after selecting on both proper motion and photometry.
The GD-1 stream is clearly visible as an overdensity of \emph{individual}
stars (the positions are not binned); this is the most pure view of the stream
to date.
This increased contrast shows that the stream extends at least another
$20^\circ$ to negative longitudes.
Furthermore, the stream reaches its highest surface density where it is the
narrowest ($\phi_1\sim-13^\circ$), which may be the location of its elusive and
fully-disrupted progenitor.
Several of the under-densities and gaps hinted at from photometric selection
(\citealt{Carlberg:2013, DeBoer:2018}) appear as striking features in this
significantly cleaner map of the GD-1 stream.
% Finally, at least two new features are visible near the stream: (1) ``the
% spur,'' stars above the main stream track (in $\phi_2$) between $-40^\circ
% \lesssim \phi_1 \lesssim -30^\circ$, and (2) ``the blob,'' stars below the main
% stream track between $-20^\circ \lesssim \phi_1 \lesssim -10^\circ$.
% We discuss each of these in more detail in \sectionname
% s~\ref{sec:results}--\ref{sec:discussion} below.
Finally, at least two new features are visible near the stream: (1) ``the
spur,'' stars above the main stream track (in $\phi_2$), and (2) ``the blob,''
stars below the main stream track, both highlighted in
\figurename~\ref{fig:selection}.
% We discuss each of these in more detail in \S\,
% \ref{sec:results}--\ref{sec:discussion} below.

@@

% We extract high-confidence stream stars from the proper motion and CMD selected
% sample by further selecting stars close to the main stream track.
% We use these main stream track stars to study the global properties of the
% stream (\S\,\ref{sec:res_global}), and find interesting substructures off of the main stream track (\S\,\ref{sec:res_gap}).

\subsection{Properties of the main GD-1 track}
\label{sec:res_global}

@@

We extract properties of the stream from the region defined above as a function
of $\phi_1$ by computing median stream properties in (overlapping) $3^\circ$ windows.
% we set the window size to $3^\circ$ in $\phi_1$, and shift the window center by
% $1^\circ$ in longitude for each subsequent measurement.
Panels (c)--(e) in \figurename~\ref{fig:track-and-model} show the
background-subtracted stream surface density and median proper motion components
extracted in this way (dark lines), with the associated empirical scatters
(shaded regions).
We estimate the local background surface density as an average of areas north
and south of the bounded stream region.
The median proper motion profiles along the stream (panels (d)--(e)) are
consistent with previous mean proper motion measurements
(\citealt{Koposov:2010}), but here the proper motion gradient is resolved for
individual stars (see, e.g., elongated overdensity in upper right panel of
\figurename~\ref{fig:selection}).

@@

Clear density variations, also apparent in the 2D positions of the sources,
manifest as sharp features in the surface density estimates along the stream.
These do not map to features in dust extinction nor are they caused by completeness variations
in the \gaia\ scanning pattern, as 99\% of our sample has at least 8 \gaia\ visibility periods, 
promising a robust estimate of the proper motions.
% First, the visible portions of the GD-1 stream are located at high Galactic
% latitudes ($b > 20^\circ$).
%, and we therefore do not expect significant dust
% extinction or variations in extinction within the stream region.
% The median $V$-band extinction over the stream track is
% $\textrm{med}\left(A_V\right) = 0.04~\textrm{mag}$ with a dispersion
% $\sigma_{A_V} \approx 0.03~\textrm{mag}$, as computed from the SFD extinction
% map (\citealt{Schlegel:1998}).
% As the stream approaches the Galactic disk ($\phi_1 < -60^\circ$),
% dust extinction becomes appreciable, and the stream stars are harder to
% select apart from the background.
% Second, 99\% of the stars in our sample have at least 8 \gaia\ visibility
% periods (median and root-variance of 14 and 3, respectively), promising a robust
% estimate of the proper motions.
% The variance in the number density of faint stars originating from the \gaia\
% scanning pattern is only $\approx 1~\textrm{deg}^{-2}$, as estimated using stars
% with proper motions $-9 < \mu_{\phi_1, \odot} < -4.5~\masyr$ and $1 <
% \mu_{\phi_2, \odot} < 4~\masyr$, far smaller than the density variations in the
% stream.
% The relative ratios between different over- and under-densities might be affected by the \gaia\ completeness at the faint end, but they
The observed density variations are therefore likely real morphological changes
along the GD-1 stream.
% as have been suggested before photometric selection alone
% (\citealt{Carlberg:2013, DeBoer:2018}).
The deep under-densities we see from the kinematically-cleaned sample correspond
to previously reported gaps in the stream (\citealt{Carlberg:2013, DeBoer:2018}).
There is also a well-defined, sharp overdensity in the surface density profile
close to $\phi_1 \approx -13^\circ$ with roughly symmetric under-densities on
either side:
this is suggestive of a progenitor system in the final stages of dissolution
(\citealt{Balbinot:2018}).
This region is also the thinnest part of the stream, as would be expected for
recently stripped material.
% that has not yet phase-mixed in multiple dimensions.

@@

@@

To compute initial conditions for the stream model, we fit an orbit to the
observed properties of the stream in a fixed Milky Way model consisting of a
disk (\citealt{Miyamoto:1975}), a bulge (\citealt{Hernquist:1990}), and a
spherical dark matter halo (\citealt{Navarro:1996}).
We set the disk scale-length parameters to match the disk model found in
\citet{Bovy:2015}; other parameters are set to produce an approximately flat
disk rotation curve with a circular velocity at the solar circle,
$v_{\textrm{circ}, \odot} \approx 230~\kms$.
%use L-BFGS-B optimization as implemented in the \package{scipy}
% \texttt{Python} package to
We maximize the likelihood of the orbit given sky track and proper motions from
this work, with mean distance and radial velocities from
\cite{Koposov:2010}.
% We do not vary the parameters of the Milky Way model in the fit because we are
% only interested in producing a comparison stream model. % of the stream properties.

\figurename~\ref{fig:orbitfit} shows the best-fitting orbit from the procedure
outlined above.
The GD-1 stream crosses the midplane at large radius, far from the bulk of the
stellar density:
the grey line in the left panel shows the radius that contains $\approx 90\%$ of
the mass, assuming a \citet{Bovy:2012} model of the stellar disk.
% an exponential radial density profile with a scale radius
% $\rm R_d = 3~\kpc$ (an intermediate value of recent measurements, e.g.,
% \citealt{Bovy:2012}).

To initialize the model stream generation, we assume that the present day
position of the progenitor system is at $\phi_1 = -13.5^\circ$, the peak of the
observed surface density profile.
We numerically integrate the backwards from this location for $4~\textrm{Gyr}$,
then forward-generate the model stream, and release stream particles at each
timestep.
We assume an initial progenitor mass of $M=10^5~\msun$, and linearly decrease
the mass of the progenitor until its full disruption at the present.
% day, where we assume $M = 0~\msun$
% (i.e. fully disrupted).
Stream properties computed from this resulting model stream are plotted in
\figurename~\ref{fig:track-and-model} as dashed (orange) lines.
The model stream qualitatively matches the stream track in all phase-space
dimensions, but has a much smoother surface density profile along the extent of
the stream.
We therefore don't expect that the observed surface density variations can be fully explained
by, e.g., epicyclic overdensities (\citealt{Kupper:2012}).
